---
title: "Chapter 9: Statistical Foundations"
author: "Nils Indreiten"
date: "21/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse,mdsr,nycflights13)
```

# Samples and Populations

## Example: Setting travel policy by sampling from the population.


Suppose you were asked to develop a travel policy for 
business travel based in NYC. The traveler has a meeting
in SFO at specified time t. The policy should be formulated
will say how much earlier than t an acceptable flight 
should arrive in order to avoid being late to the 
meeting due to a flight delay. 

For the purpose of this example we will use the 
subset of the flights in 2013 in the nycflights13 package.
The problem is to develop a policy for this year based on 
sample data. The population will be the flights filtered
for SFO:

```{r}
SF <- flights %>% 
  filter(dest == "SFO", !is.na(arr_delay))
```

We will work with just a sample of this population. Lets
set the size to be n=25:

```{r}
set.seed(101)
sf_25 <- SF %>% 
  slice_sample(n=25)
```

A simple way (albeit naive) to set up te policy is to look
for the longest flight delay and insist that travel be 
arranged to deal with the delay. 

```{r}
sf_25 %>% 
  skim(arr_delay)
```

The maximum delay is 103 minutes. Should the travel policy,
plan on arriving in SFO about 2 hours ahead? Lets look
at the complete set of flights:

```{r}
SF %>% 
  skim(arr_delay)
```

The results are different for the population relative to the
sample. The longest delay in the population was 
1007 minutes, suggesting that to avoid a meeting, should 
travel  the day before the meeting. Sure but then:

* Extra day travel, expenses including lodging, meals, 
and traveler's time.

* No guarantee there will never be a delay of more than 
10007 minutes.


A sensible travel policy will trade off small probabilities
of being late against the savings in cost and traveler's 
time. For instance, you might judge it acceptable to be 
late 2% of the time, a 98% chance of being on time.

Here's the 98th percentile of the arrivals delays in our
data sample:

```{r}
sf_25 %>% 
  summarise(q98 = quantile(arr_delay, p=0.98))
```
A delay of 84 minutes is nearly an hour and a half. So 
what? The question is really along these lines: consider a 
90-minute travel policy. how would that have worked in
achieving our intention to be late 2% of the time?

With the population data in hand, its easy to answer:

```{r}
SF %>% 
  group_by(arr_delay > 90) %>% 
  count() %>% 
  mutate(prct = n /nrow(SF)*100)
```

The 90 minute policy mark would miss 5% of the time, worse
than we intended. To hit only 2% of the time, what value
would we need to increase it to?

Lets calculate the 98th percentile of arrivals for the 
population:

```{r}
SF %>% 
  summarise(q98 = quantile(arr_delay, p = 0.98))
```

Should have been around 150. In real life tho if we do not
have a large enough sample size how can we confidently
generalise to the population as a whole? 

## Sample Statistics

### Sampling Distribution

Ultimately the goal is to figure out the reliability of 
a sample from the sample itself. If we were to collect a 
new sample from the population, how similar would the sample 
statistic on that new sample be to the same statistic 
calculated on the original sample? In other words, if we draw
many different samples form the population, each of size n,
and calculated the sample statistic on each sample, how 
similar would the sample statistic be across all samples?

Since we have the population, this is simple to figure out
using slice_simple() many times and then calculate the sample
for each trial. Here are 2 samples in which we calculate the
mean arrival delay:

```{r}
n <- 25

SF %>% 
  slice_sample(n = n) %>% 
  summarise(mean_arr_delay = mean(arr_delay))


SF %>% 
  slice_sample(n = n) %>% 
  summarise(mean_arr_delay = mean(arr_delay))
```

Perhaps its better to run many trials. The map() function from
purrr lets us automate the process. Here are the results 
from 500 (lets do 1000) trials:

```{r}
num_trials <- 1000
sf_25_means <- 1:num_trials %>% 
  map_dfr(
    ~SF %>% 
      slice_sample(n=n) %>% 
      summarise(mean_arr_delay = mean(arr_delay))
  ) %>% 
  mutate(n=n)

head(sf_25_means)
```

We now have 1000 trials, for each of which we calculated the 
mean arrival delay. Lets examine the spread:

```{r}
sf_25_means %>% 
  skim(mean_arr_delay)
```

To interpret reliability, the following vocabulary is 
necessary:

* The sample size is the number of cases in the sample, usually denoted with n. In the above, the sample size is n = 25.

* The sampling distribution is the collection of the sample statistic from all of the trials. We carried out 500 trials here, but the exact number of trials is not important so long as it is large.

* The shape of the sampling distribution is worth noting. Here it is a little skewed to the right. We can tell because in this case the mean is more than twice the median.

* The standard error is the standard deviation of the sampling distribution. It describes the width of the sampling distribution. For the trials calculating the sample mean in samples with  n=25 , the standard error is 9.22 minutes. (You can see this value in the output of skim() above, as the standard deviation of the sample means that we generated.)

* The 95% confidence interval is another way of summarizing the sampling distribution. From Figure 9.1 (left panel) you can see it is about  −16 to +20 minutes. The interval can be used to identify plausible values for the true mean arrival delay. It is calculated from the mean and standard error of the sampling distribution.

```{r}
sf_25_means %>% 
  summarise(
    x_bar = mean(mean_arr_delay),
    se = sd(mean_arr_delay)
  ) %>% 
  mutate(ci_lower = x_bar - 2 * se, # Approximately 95% obs.
         ci_upper = x_bar + 2 * se) # within 2 sd's
```

Alternatively calculate directly using a t-test:

```{r}
sf_25_means %>% 
  pull(mean_arr_delay) %>% 
  t.test()
```

An important question that statistical methods allow to 
answer is what size of sample n is needed to get a 
result with acceptable reliability. What constitutes
"acceptable" depends on the goal being accomplished.
But measuring reliability is straightforward, i.e.
it is about finding the standard error and/or
confidence interval. 

For a sample of size n=25 the range is from -17 to
57 minutes. This is important. It demonstrates the 
reliability of the sample mean for samples of arrival
delays of n=25. For n =25 the se is 9.2 minutes. 
What would happen if we used an even larger sample,
n = 100? The calculation is the same as before but 
with a different n. 

```{r}
n <- 100

sf_100_means <-  1:500 %>% 
  map_dfr(
    ~SF %>% 
      slice_sample(n=n) %>% 
      summarise(mean_arr_delay = mean(arr_delay))
  ) %>% 
  mutate(n=n)
```

```{r}
sf_25_means %>% 
  bind_rows(sf_100_means) %>% 
  ggplot(aes(x=mean_arr_delay))+
  geom_histogram(bins=30)+
  facet_grid(~n)+
  xlab("Sample mean")+
  theme_minimal()
```

## The Bootstrap

Bootstrapping is a statistical method that allows
us to approximate the sampling distribution even
without access to the population. 

Think of the the sample as the population itself.
draw many samples form the original sample.
This is _resampling_: drawing a new sample
from an existing sample. 

With resampling, sample with replacement is 
adopted, duplication is allowed. Lets consider 
the following sample (n=3) from the flights data,
setting replace=TRUE allows for ducplication.

```{r}
three_flights <- SF %>% 
  slice_sample(n=3,replace = FALSE) %>% 
  select(year,month, day, dep_time)
three_flights
```
```{r}
three_flights <- SF %>% 
  slice_sample(n=3,replace = TRUE) %>% 
  select(year,month, day, dep_time)
three_flights
```
Lets utilise bootstrapping to estimate the 
reliability of the mean arrival time calculated
on a sample of size 200:

```{r}
n <- 200
orig_sample <- SF %>% 
  slice_sample(n=n,replace = FALSE)
```

Now with this sample in hand, we can draw a 
resample of the sample size, and calculate the mean
arrival delay. 

```{r}
orig_sample %>% 
  slice_sample(n=n, replace = TRUE) %>% 
  summarise(mean_arr_delay = mean(arr_delay))
```

Repeating this process multiple times, well be 
able to see how much variation there is across
the samples:

```{r}
sf_200_bs <- 1:num_trials %>% 
  map_dfr(
    ~orig_sample %>% 
      slice_sample(n=n, replace = TRUE) %>% 
      summarise(mean_arr_delay = mean(arr_delay))
  ) %>% 
  mutate(n=n)

sf_200_bs %>% 
  skim(mean_arr_delay)
```

We can estimate that the sd is 2.3. We wouldn't 
be able to check, but because we have access to the
population data, we can. We can now compare our
bootstrap estimate to a set of hypothetical 
samples off size n=200 from the original SF 
flights.

```{r}
sf_200_pop <- 1:num_trials %>% 
  map_dfr(
    ~SF %>% 
      slice_sample(n=n, replace = TRUE) %>% 
      summarise(mean_arr_delay = mean(arr_delay))
  ) %>% 
  mutate(n=n)

sf_200_pop %>% 
  skim(mean_arr_delay)
```

Is the bootstrap standard error a good  
approximation of the standard error of the 
population? The distribution of values in the
bootstrap trials is referred to as the 
bootstrap distribution. Not the same as the 
sampling distribution. 

## Example: Setting travel policy

We can get the 98th percentile of the sample of
size n = 200, and use bootstrapping to see
how reliable that sample statistic is. The sample
itself suggests a policy of scheduling a flight
to arrive 98.16 minutes early:

```{r}
orig_sample %>% 
  summarise(q98 = quantile(arr_delay,p=0.98))
```

The reliability of that estimate can be checked
using bootstrapping:

```{r}
n <- nrow(orig_sample)
sf_200_bs <- 1:num_trials %>% 
  map_dfr(
    ~orig_sample %>% 
      slice_sample(n=n, replace = TRUE) %>% 
      summarise(q98=quantile(arr_delay, p=0.98))
  )

sf_200_bs %>% 
  skim(q98)
```

The bootstrapped standard error is about 22.7. 
Corresponding 95% confidence interval is 98.16
+- 46 minutes. A policy based on this would 
be a shot in the dark. One way to fix things 
might be to collect more data, hoping to get a 
more reliable estimate of the 98th percentile. 
Imagine we could do the work to generate a 
sample with 10,000 cases:

```{r}
set.seed(1001)

n_large <- 10000

sf_10000_bs <- SF %>% 
  slice_sample(n=n_large,replace = FALSE)

sf_200_bs <- 1:num_trials %>% 
  map_dfr(
    
    ~sf_10000_bs %>% 
      slice_sample(n=n_large, replace = TRUE) %>% 
      summarise(q98 = quantile(arr_delay, p=0.98))
  )

sf_200_bs %>% 
  skim(q98)
```
The sd is much narrower, 154 +- 8 minutes.  More 
data makes it easier to estimate, particularly in
the tails. 

## Outliers

Considering the role of outliers is important, lets say we consider
a flight delay of 7 hours (420 minutes) or more as an outlier:

```{r}
SF %>% 
  filter(arr_delay >= 420) %>% 
  select(month, day, dep_delay, arr_delay, carrier)
```

Most of the delays were in July and by VX (Virgin America). This 
suggests to potentially avoid this route. However outliers can
be misleading, they account for a tiny fraction of the flights. 
We can remove these to get a good idea of the distribution:

```{r}
SF %>% 
  filter(arr_delay < 420) %>% 
  ggplot(aes(arr_delay)) +
  geom_histogram(binwidth = 10)+
  labs(x="Arrival delay (in minutes)")
``` 

The vast majority of flights arrive without any delay or a delay of 
less than 60 minutes. Can you find a patterns that can presage 
when longer delays are likely to occur? The outlier suggests 
month and carrier might be linked to long delays. Lets see
how it plays out:

```{r}
SF %>% 
  mutate(long_delay = arr_delay > 60) %>% 
  group_by(month,long_delay) %>% 
  count() %>% 
  pivot_wider(names_from = month,
              values_from = n) %>% 
  data.frame()
```

June and July seem to be problem months:

```{r}
SF %>% 
  mutate(long_delay = arr_delay >60) %>% 
  group_by(carrier, long_delay) %>% 
  count() %>% 
  pivot_wider(names_from = carrier, values_from = n) %>% 
  data.frame()
```
Delta Airlines has good performance, this hints at a policy that might 
advise early arrival during June and July and to consider Delta when
traveling to SFO. 

## Statistical Models: Explaining variation

Another way to think about the problem at hand is that we are 
explaining part of the variation in arrival delay from flight to 
flight. Statistical modelling provides a way to relate variables
to one another. Doing so helps understand the systems we are 
studying. 

Lets put it this way:  What impact, if any, does scheduled time of
departure have on expected flight delay? Many people think that
earlier flights are less likely to be delayed. Can this be the case?

Begin by considering the time of the day:

```{r}
SF %>% 
  group_by(hour) %>% 
  count() %>% 
  pivot_wider(names_from = hour, values_from = n) %>% 
  data.frame()
```

Many flights are scheduled early to mid-morning and from late 
afternoon to early evening. None are scheduled before 5 am or
after 10 pm. Examine how arrival depends on the hour. First use 
box-and-whisker plots to show the distribution of arrival delays;
second with a kind of statistical model called a linear model
that lets us track the mean arrival delay over the course of the day:

```{r}
SF %>% 
  ggplot(aes(x=hour, y=arr_delay))+
  geom_boxplot(alpha=0.1, aes(group = hour))+
  geom_smooth(method = "lm", se = TRUE) +
  xlab("Scheduled hour of departure")+
  ylab("Arrival delay (minutes)")+
  coord_cartesian(ylim = c(-30, 120))+
  theme_classic()
```

The trend line is created via a regression model:

```{r}
modd1 <- lm(arr_delay ~ hour, data = SF)
broom::tidy(modd1)
```

 This suggests that arrival delay is predicted to be about 2 minutes
 higher per hour. Over the 15 hours of flights, this leads to a 
 30 minute increase in arrival delay comparing flights at the end
 of the day to the beginning of the day. Broom also
 calculates the standard error: 0.09 minutes per hour. Stated
 as a 95% confidence interval, this model indicates that we are 
 95% confident that the true arrival delay increase by 2 +- 0.18
 minutes per hour.
 
 Can we do better? What additional factors might help to explain flight
 delays? departure airport, carrier, month of the year, and day of the
 week. some wrangling is necessary: extract day of the week from year,
 month, and day of the month. Also create a variable season that 
 summarises what we already know about the month:June and July 
 are months with long delays. Used as explanatory variables:
 
```{r}
library(lubridate)

SF <- SF %>% 
  mutate(
    day = as.Date(time_hour),
    dow = as.character(wday(day,label = TRUE)),
    season = ifelse(month %in% 6:7, "Summer", "Other Month")
  )
```
 
Build a model that includes variables we want to use to explain 
arrival delay:

```{r}
mod2 <- lm(arr_delay ~ hour + origin + carrier + season + dow, 
           data = SF)
broom::tidy(mod2)
```

Estimate denotes how many minutes should be added to the average 
delay. Linear models describe how the mean of the response variable
varies with the explanatory variables. 

## Confounding and accounting for other factors

The causation correlation problem, Lets consider the example
of confounding using observational data on average teacher salaries
and average total SAT scores for each State. We can fit a model
to see how the realtionship unfolds:

```{r}
SAT_2010 <- SAT_2010 %>% 
  mutate(Salary = salary/1000)

SAT_plot <- 
  SAT_2010 %>% 
  ggplot(aes(Salary, total))+
  geom_point()+
  geom_smooth(method="lm")+
  ylab("Average total score on SAT")+
  xlab("Average teacher salary (Thousands of USD)")

SAT_plot
```

```{r}
SAT_mod1 <- lm(total~Salary, data = SAT_2010)
broom::tidy(SAT_mod1)
```

There is another factor to consider. The percentage of students who
take the SAT in each state varies dramatically. Lets quickly
get a sense of the distribution:

```{r}
SAT_2010 %>% 
  skim(sat_pct)
```

Lets create a variable referred to as SAT_grp that divides the 
states into two groups:

```{r}
SAT_2010 %>% 
  mutate(SAT_group = ifelse(sat_pct <= 27, "Low","High")) -> SAT_2010

SAT_2010 %>% 
  group_by(SAT_group) %>% 
  count()
```
Lets visualise the previous plot but this time broken down by 
SAT_group:

```{r}
SAT_plot %+% SAT_2010 +
  aes(color= SAT_group)+
  scale_color_brewer("% taking\nthe SAT", palette = "Set1")
```

We can derive the coefficients of the linear model fit to the 
two separate groups:

```{r}
SAT_2010 %>% 
  group_by(SAT_group) %>% 
  group_modify(~broom::tidy(lm(total ~Salary, data=.x)))
```

For each of the groups, notice the possibility here to fit models
to groups without having to use nest(). average teacher salary
is positively associated with average SAT score. However when 
collapsing over this variable, average teacher salary is 
negatively associated with average SAT score. This form
of compounding is a quantitative version of Simpson's paradox
and arises in many situations. The summary is:

* Among state with low percentage taking the SAT, teacher salaries
and SAT scores are positively associated. 

* Among states with a high % of taking the SAT, salaries and
SAT scores are positively associated. 

* Among all states, salaries and SAT scores are negatively associated. 

Addressing this issue is straightforward if the confounding 
variables are measure. Stratification (as done above) is one option.
Multiple regression is another. Lets add sat_pct as an additional 
predictor into the regression model:

```{r}
SAT_mod2 <- lm(total ~ Salary + sat_pct, data = SAT_2010)
broom::tidy(SAT_mod2)
```

We now see that the slope for Salary is positive and statistically 
significant when we control for sat_pct. This is consistent with the
results when the model was stratified by SAT_group. 

We still can't conclude that salaries cause improvements in SAT scores.
However, the associations after confounding are likely more reliable. 

## Exercises:

# Problem 1:

3 sd.

# Problem 2:

Calculate end interpret a 95% confidence interval of the mean age
of mothers from Gestation from the mosaicData package:

```{r}
library(mosaicData)

Gestation %>% 
  skim(age)

```

The standard deviation for the age in the Gestation dataset is 
5.78, which therefore indicated that 95% of the age in the population
falls within ~ +- 10 years. 



# Problem 3 

Use the bootstrap to generate and interpret a 95% confidence interval
for the median age of mothers for the Gestation data:

```{r}
n <- 100
num_trials <- 10000

Gestation_200_bs <- 1:num_trials %>% 
  map_dfr(
    ~Gestation %>% 
      slice_sample(n=n, replace = TRUE) %>% 
      summarise(median_age = median(age))
  ) %>% 
  mutate(n=n)

Gestation_200_bs %>% 
  skim(median_age)

```


## Problem 4

```{r}
mosaicData::Gestation ->Gestation 
library(rsample)

bootstrap_mod <- reg_intervals(wt~age , 
                               data = Gestation,
                               times=500)
bootstrap_mod
```


## Problem 5

Describe the association between smoking status and mortality
in the Whickham data set from MosaicData:

```{r}
Whickham <- mosaicData::Whickham
```

Lets explore the data first:

```{r}
Whickham %>% 
  ggplot(aes(age))+
  geom_histogram(bins = 15)+
  facet_wrap(~smoker)
```

Change the variables to factor:

```{r}
Whickham %>% 
  mutate_if(is.character,as.factor) -> Whickham
```

Fit a logistic regression model:

```{r}
glm(outcome ~smoker + age,family = "binomial",data = Whicham)->W_model
# Visualise the model
library(visreg)
visreg(W_model,
       "age",
       by="smoker",
       gg=TRUE,
       scale = "response")+
  labs(y="Prob(Death)",
       x="Age",
       title = "Relationship of age and death",
       subtitle = "controlling for smoker",
       caption = "source: MosaicData::Whickham")+
  theme_minimal()
```



















